From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-04 Pi+/paper/ufin.tex
Text about homotopies and equivalences

The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
invertible cells for the left and right units of composition, and associativity. The higher cells come with their own
coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category
  in appendix.}

@@

\begin{gather*}
  \begin{aligned}
    \term{sym}   & : (x \id_{A} y) \to (y \id_{A} x)                   \\
    \term{\comp} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
  \end{aligned}
  \qquad
  \begin{aligned}
    \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w)    \\
                 & \to (p \comp q) \comp r \id p \comp (q \comp r)             \\
    \term{invr}  & : (p : x \id_{A} y) \to p \comp \term{sym}(p) \id \refl_{x}
  \end{aligned}
\end{gather*}

@@

Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences of fibers.

@@

\todo{Diagram.}

@@

\subsubsection{Univalence}

Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or,
the following function is an equivalence.

\[
  \ua : A \id_{\UU} B \to A \eqv B
\]

\todo{Revise.}

Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.

>>>>

The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$. In a weak category, equational
axioms of the category are replaced by higher invertible cells. For example, there are higher invertible cells for the
left and right units of composition, and associativity. The higher cells come with their own coherence data as well,
such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category in appendix.}

@@

\begin{gather*}
  \begin{aligned}
    \term{\inv{\blank}}      & : (x \id_{A} y) \to (y \id_{A} x)                   \\
    \term{\blank\comp\blank} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
  \end{aligned}
  \qquad
  \begin{aligned}
    \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w) \\
                 & \to (p \comp q) \comp r \id p \comp (q \comp r)          \\
    \term{invr}  & : (p : x \id_{A} y) \to p \comp \inv{p} \id \refl_{x}
  \end{aligned}
\end{gather*}

A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
function extensionality. An equivalent between types $A \eqv B$ is given by a pair of functions between them which
compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$. This is equivalent to the
identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

Further, using the groupoid structure of $A$, we can show that $\transport{P}{p}{\blank}$ and
$\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name the map $\tptEqv{P}$ which lifts paths to equivalences
of fibers.

@@

\todo{Diagram.}
The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent fibrations.

@@

The use of the word \emph{univalent} here is a reference to Voevodsky's \emph{univalence} principle. Indeed, univalence
characterises paths in the universe as equivalences between types, which follows from the canonical fibration $\idfunc :
\UU \to \UU$ being univalent.
